---
title: "Regression Models - Course Project"
geometry: margin=2cm
output: html_document
fontsize: 10pt
---
```{r echo=FALSE,message=FALSE,warning=FALSE,error=FALSE,cache=TRUE}
require(GGally);require(car);
data(mtcars);
all<-lm(mpg~.,mtcars)
wthp<-lm(mpg~wt+hp,mtcars)
#wthpqsec<-lm(mpg~wt+hp+qsec,mtcars)
wthpam<-lm(mpg~wt+hp+am,mtcars)
wthpqsecam<-lm(mpg~wt+hp+qsec+am,mtcars)
wthpqsecamdspl<-lm(mpg~wt+hp+qsec+am+disp,mtcars)
models<-list("all"=all,
             "wthp"=wthp,
             "wthpam"=wthpam,
             "wthpqsecam"=wthpqsecam,
             "wthpqsecamdspl"=wthpqsecamdspl)

modeladjrsquare<-sapply(models,function(x){summary(x)$adj.r.squared})
names(modeladjrsquare)<-names(models)
```

## Executive Summary
Analysis to answer the following two questions based on a motor vehicle data set.  
1) "Is an automatic or manual transmission better for MPG"  
2) "How different is the MPG between automatic manual transmission?"  
Several regression models were evaluated, compared for accuracy and used to answer the above questions.

## Results
Detailed analysis with the dataset revealed we cannot find a statistically significant effect of 
transmission type on gas mileage(mpg).

The slope coefficient of am is 2.08 with a **p value of 0.14** in a model using wt,hp and am.
This p value is not enough to to reject the null hypothesis that transmission has no effect on mpg.
If the p-value had been found to be lower than our significance level this slope of 2.08
would represent improvement in gas mileage of manual transmission over automatic transmission vehicles,
answering question 2. However this slope currently does not mean anything significant.

Weight and horsepower were found to be the two most significant factors in estimating mpg 
with p values close to zero.

## Description of analysis done
Initially a multivariable regression was done with all available data to serve as a baseline.
This analysis is not ideal as it does not address multicollinearity, and includes several variables that ought not to affect mpg.  

###Model Selection
From studying several semesters of automobile engineering during B.Tech(Mechanical) I can classify the factors as YES/NO/MAYBE  [Expert Opinion]

NO: Axle ratios, Gear/Carburetor/Cylinder count, V/Straight engine layout, will not affect mpg.  
YES: Weight and horsepower(hp) definitely affect mpg. To produce more power more gas is burnt. 
Energy expended to move a Weight is directly proportional to its Weight.  
MAYBE:  Quarter mile time  can be considered but would be expected to show collinearity with hp. 
Engine displacement(size) only has an indirect effect on power and not a very linear one. However it was evaluated.

####Exploratory Analysis
A pairs plot of the factors considered possibly relevant is shown in [appendix](#appendix)  
This shows a high negative correlation between mpg vs weight, displacement & hp.
Displacement is also highly correlated with weight[0.89] and hp[0.79]

Four additional models were constructed. The model names reflect the factors they include  
**wthp | wthpam | wthpqsecam | wthpqsecamdspl  **

For instance wthp uses wt and hp, wthpam uses wt,hp and am and so on.

####Adjusted R Square of different models
```{r adjrsqcomp, echo=FALSE,cache=TRUE}
modeladjrsquare
```
Adjusted r square values showed  improvement in using wt&hp over using all variables. 
Adding in am and qsec also improved adj.r.square but adding displ barely improved it.

p value for the F statistic of the fitted models to examine model significance
```{r Fstat,echo=FALSE,message=F,warning=F,error=FALSE,cache=TRUE}
sapply(models,function(x){f<-summary(x)$fstatistic;
                          p<-pf(f[1],f[2],f[3],lower.tail=F)
                          attributes(p)<-NULL
                          p})
```
The model with just wt and hp seems most significant from its lowest p value. 
This implies that Weight and Horsepower seem to explain most of the variation in mpg. 

####Evaluate collinearity using Variance Inflation Factors
```{r echo=FALSE,cache=TRUE}
vif(all)
```
cyl, disp and weight showed the highest variance inflation factors indicating possible collinearity. 
This also makes intuitive sense since larger displacement engines often have more cylinders, 
are heavier and are typically installed in heavier vehicles. In the model with all variables 
cyl and displ coefficients have p values of 0.91 and 0.46, indicating they are not significant. 
Remove the models using cyl and dspl from consideration.

```{r anova, echo=FALSE,cache=TRUE}
models$wthpqsecamdspl<-NULL
anova(wthp,wthpam,wthpqsecam)
```

From the anova results, p value for adding am to wt+hp [0.12590] is higher than our significance level - 
adding am to a model having wt+hp does not improve the model significantly. Adding qsec to the 
wthpam model did not improve it either [p value 0.076].

####Model coefficients
```{r summarycoeffs, echo=FALSE,cache=TRUE}
summary(wthpam)$coefficients
```
The p value for am is 0.14 which is much much higher than permissible(0.01 or even 0.05).
**Based on the p-value and anova results we conclude that the trasmission type has no significant effect on gas mileage.**

wt and hp have a significant effect on mileage a model having just wt and hp is finalized.

Model [residuals](#residuals) do not show any pattern in their scatter which is good.
The residuals do show some skew in their [distribution](#distribution)
but for such a small sample size we can ignore this.

## Appendix

```{r echo=FALSE,warning=FALSE,error=F, fig.width=10,fig.height=15,cache=TRUE}
ggpairs(c(1,3,4,6,7,9),data=mtcars,lower=list(continous="smooth",params=c(method="loess")))
```

### <a name="residuals"/> Residuals of model with weight and horsepower

```{r residualplot, echo=FALSE,warning=FALSE,error=F, fig.width=10,fig.height=15,cache=TRUE}
par(mfrow=c(2,2))
plot(wthp)
```

### <a name="distribution"/>Distribution of residuals
```{r residualshist, echo=F,warning=F,error=F,fig.width=10,cache=TRUE}
hist(wthp$residuals,col="blue",xlab="Residuals",main="Residuals of model with wt and hp")
```
